---
title: "Project Proposal"
author: "Anvita, Matthew, Brynn, Pinaki"
subtitle: "Due November 16 at 11:59pm"
format: pdf
editor: visual
---

#### Load Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(dplyr)
```

# Dataset 1 (top choice)

**Data source:** Current Population Survey (CPS) – Merged Outgoing Rotation Group (MORG) Earnings Microdata

One of the dataset from the ‘[The National Bureau of Economic Research](https://www.nber.org/research/data?page=1&perPage=50)’

**Source link:** [https://www.nber.org/research/data/current-population-survey-cps-merged-outgoing-rotation-group-earnings-data](https://www.nber.org/research/data/current-population-survey-cps-merged-outgoing-rotation-group-earnings-data?utm_source=chatgpt.com)

**Brief description:**

The CPS MORG dataset contains detailed monthly labor market microdata collected jointly by the U.S. Census Bureau and the Bureau of Labor Statistics. Each annual MORG file includes around **300,000+ individual-level observations**. Each row represents an adult respondent who was in the “outgoing rotation group” for that survey month.

The dataset includes a rich mix of variables:

-   **Demographics:** age, gender, race, education level, marital status

-   **Geography:** state, region, metro status

-   **Labor & earnings:** hourly wage, weekly earnings, hours worked, union membership, occupation, industry

-   **Weights:** person-level and earnings weights

For a class project, a clean subset can be created by selecting full-time employed adults (ages 25–64) and choosing 10–15 variables with a mix of numeric (e.g., wage, hours, age) and categorical (e.g., gender, education group, union status, industry).

**Research Question 1**

**Research question 1: How does union membership relate to hourly wages after accounting for education, experience (proxied by age), gender, and industry?**

-   Outcome variable (include the name/description and type of variable):
    -   **Hourly Wage**
        -   **Type:** Numeric (continuous)

        -   **Description:** Reported or computed hourly earnings for the respondent.
-   Main predictor variable
    -   **Union Membership (UNION)**
        -   **Type:** Binary categorical

        -   **Description:** Indicates whether the worker is a union member or covered by a union contract.
-   Other potential predictors
    -   **Education group (categorical)**
    -   **Age (numeric)**
    -   **Gender (categorical)**
    -   **Industry (categorical)**
    -   **Race (categorical)**
    -   **Variable Type**
-   **Numeric outcome**, categorical + numeric predictors.

**Research question 2: Does the gender wage gap persist after controlling for age, education, occupation, industry, and hours worked?**

-   Outcome variable (include the name/description and type of variable):
    -   **Hourly Wage (continuous), same as above**
-   Main predictor variable
    -   **Gender (Male/Female)**
        -   **Type:** Categorical

        -   **Description:** Respondent’s self-reported gender.
-   **Other potential predictors**
    -   **Usual hours worked (numeric)**
    -   **Education level (categorical)**
    -   **Occupation group (categorical)**
    -   **Industry (categorical)**
    -   **Age (numeric)**
    -   **Race/ethnicity (categorical)**
-   **Interaction term idea**
    -   **Gender × Occupation group**\
        To test whether wage differences vary depending on the type of occupation (e.g., management vs. service work).

**Load the data and provide a `glimpse()`**:

```{r data-1, warning=FALSE}

```

**Exploratory Plots:**

```{r plots-1}

```

# Dataset 2

**Data source:** United Nations Data, Gini index, CPIA policy and institutions for environmental sustainability rating (1=low to 6=high), Life expectancy at birth for both sexes combined (years), GDP per capita , PPP (current international \$)

**Brief description:** The UN collects data and a plethora of metrics for countries around the world. The Gini index is a measure of wealth distribution, with 0 representing perfect equality and 100 representing one person owning all wealth. The CPIA policy and institutions for environmental sustainability rating (1=low to 6=high) is a country's policies and institutional frameworks for environmental protection and sustainable resource management. Life expectancy at birth is how long a given person is expected to live at birth. GDP per capita is the Gross Domestic Product of a country divided by its population, a measure of economic output.

**Research question 1:** How is life expectancy at birth related to Gini index controlling for GDP per capita?

-   Outcome variable (include the name/description and type of variable): Life expectancy at birth (years), continuous quantitative.
-   We could include an interaction term between Gini Index and Region. The effect of income inequality on life expectancy might differ by geographic region. For example, high inequality in Latin American countries might affect health outcomes differently than high inequality in Sub-Saharan African countries.

**Research question 2:** How is the sustainability of a country's government related to Gini index controlling for GDP per capita?

-   Outcome variable (include the name/description and type of variable): CPIA policy and institutions for environmental sustainability rating (1=low to 6=high), discrete quantitative / ordinal categorical
-   We could include an interaction term between Gini Index and Year. As global awareness of climate change has increased, countries with different inequality levels might respond differently to pressure for environmental policy reform over time. The effect of Gini index on sustainability in 2010 might be different than in 2020.

**Load the data and provide a `glimpse()`**:

```{r data-2, warning=FALSE}
life_exp <- read.csv("UNLifeExp.csv")
glimpse(life_exp)
gdp_per_cap <- read.csv("UNGDP.csv")
glimpse(gdp_per_cap)
gini <- read.csv("UNGini.csv")
glimpse(gini)
sustainability <- read.csv("UNSust.csv")
glimpse(sustainability)
```

**Exploratory Plots:**

```{r plots-2}

# Clean Life Expectancy data
life_exp_clean <- life_exp |>
  filter(
    !str_detect(Country.or.Area, "region|Region|World|WORLD|income|Income|developed|Developed"),
    Variant == "Medium",
    Year.s. >= 1990 & Year.s. <= 2023
  ) |>
  select(Country = Country.or.Area, Year = Year.s., LifeExp = Value) |>
  filter(!is.na(LifeExp))

# Clean GDP per capita data
gdp_clean <- gdp_per_cap |>
  filter(!str_detect(Country.or.Area, "region|Region|World|WORLD|income|Income|developed|Developed")) |>
  filter(str_detect(Year, "^\\d+$")) |>  # Keep only numeric years
  mutate(Year = as.integer(Year)) |>
  select(Country = Country.or.Area, Year, GDP = Value) |>
  filter(!is.na(GDP))

# Clean Gini index data
gini_clean <- gini |>
  filter(!str_detect(Country.or.Area, "region|Region|World|WORLD|income|Income|developed|Developed")) |>
  filter(str_detect(Year, "^\\d+$")) |>  # Keep only numeric years
  mutate(Year = as.integer(Year)) |>
  select(Country = Country.or.Area, Year, Gini = Value) |>
  filter(!is.na(Gini))

# Clean Sustainability data
sust_clean <- sustainability |>
  filter(!str_detect(Country.or.Area, "region|Region|World|WORLD|income|Income|developed|Developed")) |>
  select(Country = Country.or.Area, Year, Sustainability = Value) |>
  filter(!is.na(Sustainability))
```

```{r}

# Merge datasets and use most recent year available for each country
# Merge for Research Question 1 (Life Expectancy)
rq1_data <- life_exp_clean |>
  inner_join(gini_clean, by = c("Country", "Year")) |>
  inner_join(gdp_clean, by = c("Country", "Year")) |>
  group_by(Country) |>
  slice_max(Year, n = 1) |>
  ungroup()

# Merge for Research Question 2 (Sustainability)
rq2_data <- sust_clean |>
  inner_join(gini_clean, by = c("Country", "Year")) |>
  inner_join(gdp_clean, by = c("Country", "Year")) |>
  group_by(Country) |>
  slice_max(Year, n = 1) |>
  ungroup()
```

```{r}
# PLOT 1: Distribution of Life Expectancy (Outcome for RQ1)
ggplot(rq1_data, aes(x = LifeExp)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Life Expectancy at Birth",
    x = "Life Expectancy (years)",
    y = "Number of Countries"
  ) +
  theme_minimal()

# PLOT 2: Life Expectancy vs Gini Index (Primary relationship for RQ1)
ggplot(rq1_data, aes(x = Gini, y = LifeExp)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Life Expectancy vs. Gini Index",
    x = "Gini Index (0 = perfect equality, 100 = perfect inequality)",
    y = "Life Expectancy at Birth (years)"
  ) +
  theme_minimal()
```

```{r}
# PLOT 3: Distribution of Sustainability Rating (Outcome for RQ2)
ggplot(rq2_data, aes(x = Sustainability)) +
  geom_histogram(binwidth = 0.5, fill = "forestgreen", color = "white") +
  labs(
    title = "Distribution of Environmental Sustainability Rating",
    x = "CPIA Sustainability Rating (1 = low, 6 = high)",
    y = "Number of Countries"
  ) +
  scale_x_continuous(breaks = 1:6) +
  theme_minimal()

# PLOT 4: Sustainability vs Gini Index (Primary relationship for RQ2)
ggplot(rq2_data, aes(x = Gini, y = Sustainability)) +
  geom_point(alpha = 0.7, color = "forestgreen") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Environmental Sustainability Rating vs. Gini Index",
    x = "Gini Index (0 = perfect equality, 100 = perfect inequality)",
    y = "CPIA Sustainability Rating (1-6)"
  ) +
  theme_minimal()
```

# Team Charter

#### When will you meet as a team to work on the project components? Will these meetings be held in person or virtually?

We will aim to meet 4 times a week (3 times in person and once virtually) for an hour each. This will be to delegate any individual work, answer questions anyone may have, and to work together to analyze our results. We will add additional meetings as needed before the deadline.

#### What is your group policy on missing team meetings (e.g., how much advance notice should be provided)?

Team members should provide a 24 hour notice if they cannot attend a meeting and should just notify as soon as possibly if an emergency comes up. They should catch up with one member of the group to see what was missed and text in the group chat their update on their portion of the project.

#### How will your team communicate (email, Slack, text messages)? What is your policy on appropriate response time (within a certain number of hours? Nights/weekends?)?

We have a Whatsapp message group chat where will communicate. Team members should always reply as soon as possible (within a few hours during the day).
