---
title: "Project Proposal"
author: "Anvita, Matthew, Brynn, Pinaki"
subtitle: "Due November 16 at 11:59pm"
format: pdf
editor: visual
---

#### Load Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(dplyr)
```

# Dataset 1 (top choice)

**Data source:** Current Population Survey (CPS) – Merged Outgoing Rotation Group (MORG) Earnings Microdata

One of the dataset from the ‘[The National Bureau of Economic Research](https://www.nber.org/research/data?page=1&perPage=50)’

**Source link:** [https://www.nber.org/research/data/current-population-survey-cps-merged-outgoing-rotation-group-earnings-data](https://www.nber.org/research/data/current-population-survey-cps-merged-outgoing-rotation-group-earnings-data?utm_source=chatgpt.com)

**Brief description:**

The CPS MORG dataset contains detailed monthly labor market microdata collected jointly by the U.S. Census Bureau and the Bureau of Labor Statistics. Each annual MORG file includes around **300,000+ individual-level observations**. Each row represents an adult respondent who was in the “outgoing rotation group” for that survey month.

The dataset includes a rich mix of variables:

-   **Demographics:** age, gender, race, education level, marital status

-   **Geography:** state, region, metro status

-   **Labor & earnings:** hourly wage, weekly earnings, hours worked, union membership, occupation, industry

-   **Weights:** person-level and earnings weights

For a class project, a clean subset can be created by selecting full-time employed adults (ages 25–64) and choosing 10–15 variables with a mix of numeric (e.g., wage, hours, age) and categorical (e.g., gender, education group, union status, industry).

**Research Question 1**

**Research question 1: How does union membership relate to hourly wages after accounting for education, experience (proxied by age), gender, and industry?**

-   Outcome variable (include the name/description and type of variable):
    -   **Hourly Wage**
        -   **Type:** Numeric (continuous)

        -   **Description:** Reported or computed hourly earnings for the respondent.
-   Main predictor variable
    -   **Union Membership (UNION)**
        -   **Type:** Binary categorical

        -   **Description:** Indicates whether the worker is a union member or covered by a union contract.
-   Other potential predictors
    -   **Education group (categorical)**
    -   **Age (numeric)**
    -   **Gender (categorical)**
    -   **Industry (categorical)**
    -   **Race (categorical)**
    -   **Variable Type**
-   **Numeric outcome**, categorical + numeric predictors.

**Research question 2: Does the gender wage gap persist after controlling for age, education, occupation, industry, and hours worked?**

-   Outcome variable (include the name/description and type of variable):
    -   **Hourly Wage (continuous), same as above**
-   Main predictor variable
    -   **Gender (Male/Female)**
        -   **Type:** Categorical

        -   **Description:** Respondent’s self-reported gender.
-   **Other potential predictors**
    -   **Usual hours worked (numeric)**
    -   **Education level (categorical)**
    -   **Occupation group (categorical)**
    -   **Industry (categorical)**
    -   **Age (numeric)**
    -   **Race/ethnicity (categorical)**
-   **Interaction term idea**
    -   **Gender × Occupation group**\
        To test whether wage differences vary depending on the type of occupation (e.g., management vs. service work).

**Load the data and provide a `glimpse()`**:

```{r data-1, warning=FALSE}

```

**Exploratory Plots:**

```{r plots-1}

```

# Dataset 2

**Data source:** United Nations Data, Gini index, CPIA policy and institutions for environmental sustainability rating (1=low to 6=high), Life expectancy at birth for both sexes combined (years), GDP per capita , PPP (current international \$)

**Brief description:** The UN collects data and a plethora of metrics for countries around the world. The Gini index is a measure of wealth distribution, with 0 representing perfect equality and 100 representing one person owning all wealth. The CPIA policy and institutions for environmental sustainability rating (1=low to 6=high) is a country's policies and institutional frameworks for environmental protection and sustainable resource management. Life expectancy at birth is how long a given person is expected to live at birth. GDP per capita is the Gross Domestic Product of a country divided by its population, a measure of economic output.

**Research question 1:** How is life expectancy at birth related to Gini index controlling for GDP per capita.

-   Outcome variable (include the name/description and type of variable): Life expectancy at birth (years), continuous quantitative.

**Research question 2:** How is the sustainability of a country's government related to Gini index controlling for GDP per capita.

-   Outcome variable (include the name/description and type of variable): CPIA policy and institutions for environmental sustainability rating (1=low to 6=high), discrete quantitative / ordinal categorical

**Load the data and provide a `glimpse()`**:

```{r data-2, warning=FALSE}
life_exp <- read.csv("UNLifeExp.csv")
glimpse(life_exp)
gdp_per_cap <- read.csv("UNGDP.csv")
glimpse(gdp_per_cap)
gini <- read.csv("UNGini.csv")
glimpse(gini)
sustainability <- read.csv("UNSust.csv")
glimpse(sustainability)
```

**Exploratory Plots:**

```{r plots-2}
#Cut out regions and only include cuntries for each dataset
#Cut out NAs
#For life expectancy, look into why future years are included and cut out variant column
#For GDP, cut out footnotes column
#For Gini, cut out footnotes column
#For Sustainability, cut out footnotes column and factor values to be [1, 2), [2, 3), etc.
```

# 

# Dataset 3 (optional)

**Data source:**

**Brief description:**

**Research question 1:**

-   Outcome variable (include the name/description and type of variable):

**Research question 2:**

-   Outcome variable (include the name/description and type of variable):

**Load the data and provide a `glimpse()`**:

```{r data-3, warning=FALSE}

```

**Exploratory Plots:**

```{r plots-3}

```

# 

# Team Charter

#### When will you meet as a team to work on the project components? Will these meetings be held in person or virtually?

#### What is your group policy on missing team meetings (e.g., how much advance notice should be provided)?

#### How will your team communicate (email, Slack, text messages)? What is your policy on appropriate response time (within a certain number of hours? Nights/weekends?)?

#### 
