---
title: "Preprocessing for Dataset1: CPS MORG 2023"
format: pdf
editor: visual
---

## Preprocessing Steps

This part of the code reads the raw CPS MORG 2023 file (`morg23.dta`),

Applies all preprocessing and scoping rules, and saves a cleaned dataset that will be used in the proposal QMD.

## Steps involved:

1.  Read the complete file as downloaded from the source- morg23.dta (243484 obs of 98 varaiables)
2.  Identifies the relevant columns that we want to take for the project
3.  Subsets the data to keep only those columns
4.  Filter & Transform the data basis the following conditions:
    1.  South Atlantic States- filter
    2.  Working Age Adults- filter
    3.  Having positive hours and weekly earnings- filter
    4.  Drops all self-employed and unpaid workers- filter
    5.  Calculates hourly wages- transform
    6.  Excludes outliers (where the hourly wage is less than 5 or more than 150)- outlier treatment
    7.  Factor the following variables:
        1.  Hourly wage- 1 if hourly wage \> 40 and 0 if it is less than 40
        2.  Sex- 1 if Male and 0 if Female
        3.  `educ` (from `grade92`) - Ex: 31- High School, 40- Some College
        4.  race - Ex: 1- White, 2-Black
        5.  unionme (Union Membership Flag)- 1- Member, 0- NonMember
        6.  Full time or Part time status- ftpt94- Ex: 1-Full time, 2- Parttime, 3-Varying Hours
        7.  Class of Worker (class94)
        8.  Labor force status (lfsr94)
    8.  Create a sample set of 5000 rows as the data post filtering was still 16000 rows
    9.  Writing the data into the folder to be used for EDA and further project analysis

Raw file lives at: `data/morg23.dta`

Output files will be written to:\
`data/cps23_southatlantic_project.rds`\
`data/cps23_southatlantic_project.csv`

```{r}
library(haven)
library(dplyr)
library(ggplot2)
```

You can add options to executable code like this

```{r}
# 1. Loading raw NBER MORG data for 2023
cps_raw <- read_dta("data/morg23.dta")

# 2. Inspecting variable names
names(cps_raw)

# 3. Selecting specific variables for the project dataset
cps_sel <- cps_raw %>%
  select(
    year,
    intmonth,
    stfips,      # state FIPS
    age,
    sex,
    race,
    grade92,
    uhourse,     
    earnwke,
    class94,
    unionmme,
    lfsr94,
    ftpt94,
    earnwt,
    weight
  ) %>%
  rename(
    month = intmonth,
    state = stfips,
    hours = uhourse
  )

# 4.Applying filters --> South Atlantic FIPS states:
# 10 DE, 11 DC, 12 FL, 13 GA, 24 MD, 37 NC, 45 SC, 51 VA, 54 WV

cps23 <- cps_sel %>%
  # working-age adults
  filter(!is.na(age),
         age >= 25, age <= 64) %>%
  # positive hours and weekly earnings
  filter(!is.na(hours), hours > 0,
         !is.na(earnwke), earnwke > 0) %>%
  # South Atlantic states only
  filter(state %in% c(10, 11, 12, 13, 24, 37, 45, 51, 54)) %>%
  # drop self-employed / unpaid (keep 1–4 = gov + private wage/salary)
  filter(class94 %in% c(1, 2, 3, 4)) %>%
  # hourly wage from weekly earnings / usual hours
  mutate(
    hrwage = as.numeric(earnwke) / as.numeric(hours)
  ) %>%
  # trim outliers
  filter(!is.na(hrwage),
         hrwage >= 5, hrwage <= 150) %>%
  # derived vars + factors
  mutate(
    high_earner = if_else(hrwage >= 40, 1L, 0L),
    sex  = factor(sex,  levels = c(1, 2),
                  labels = c("Male", "Female")),
    educ = factor(grade92),
    race = factor(race),
    unionmme = factor(unionmme),
    ftpt94   = factor(ftpt94),
    class94  = factor(class94),
    lfsr94   = factor(lfsr94)
  ) %>%
  select(
    year, month, state,
    age, sex, race, educ,
    hours, hrwage, high_earner,
    earnwke,
    unionmme, lfsr94, ftpt94, class94,
    earnwt, weight
  )

# 5. Taking a sample of 5000 rows

set.seed(702)

if (nrow(cps23) > 5000) {
  cps23_proj <- cps23 %>% slice_sample(n = 5000)
} else {
  cps23_proj <- cps23
}

nrow(cps23_proj)

# 6 Writing the samples back for use in the Proposal QMD
write.csv(
  cps23_proj,
  "data/cps23_southatlantic_project.csv",
  row.names = FALSE
)

saveRDS(
  cps23_proj,
  "data/cps23_southatlantic_project.rds"
)
```
